### 2019-10-31 README.Rmd
Made minor text changes to README

In 2012 KEGG released its first application programming interface (API), and has been
adding features and functionality ever since. There are benefits to using an API.
First, API's like KEGG's allow users to perform customized analyses with the most
up-to-date versions of the data contained in the database. In addition, accessing the
KEGG API is very easy using statistical programming tools like R or Python and
integrating data retrieval into user's code makes the program reproducible. To 
further enforce reproducibilty *diffEnrich* adds a date and KEGG release tag to
all data files it generates from accessing the API. For update histories and release 
notes for the KEGG REST API please go [here](https://www.kegg.jp/kegg/rest/).

@@

**Figure 1.** diffEnrich Analysis pipeline. Functions within the diffEnrich package are represented by blue rectangles. The data that must be provided by the user is represented by the purple ovals. Data objects generated by a function in diffEnrich are represented by red ovals. The external call of the get_kegg function to the KEGG REST API is represented in yellow.

@@

Often high throughput omics studies include a functional enrichment analysis to glean biological insight from a list of candidate genes, proteins, metabolites, etc. Functional enrichment examines whether the number of genes in the list associated with a biological function or particular pathway is more than would be expected by chance.  As an example, enrichment of a particular pathway among a list of genes that are differentially expressed after an experimental manipulation may indicate that the pathway has been altered by that manipulation. This analysis is rather straight forward and many solutions have been offered (e.g., [Haung et al., 2009](https://www.ncbi.nlm.nih.gov/pubmed/19033363); [Kuleshov et al., 2016](https://www.ncbi.nlm.nih.gov/pubmed/27141961); [Liao et al., 2019](https://www.ncbi.nlm.nih.gov/pubmed/31114916); [Subramanian et al., 2005](https://www.pnas.org/content/102/43/15545)) as well a wide variety of databases have been used to define these pathways (e.g., [Kanehisa and Goto, 2000](https://www.ncbi.nlm.nih.gov/pubmed/10592173)) and ontologies (e.g., [Ashburn et al., 2000](https://www.ncbi.nlm.nih.gov/pubmed/10802651)).

@@

### Step 1: Collect and store pathways from KEGG API
First we will use the *get_kegg* function to access the KEGG REST API and download the data sets required to perform 
our downstream analysis. This function takes two arguments. The first, 'species' is required. Currently, *diffEnrich* supports three species, and the argument is a character string using the KEGG
code [https://www.pnas.org/content/suppl/2008/09/11/0806162105.DCSupplemental/ST1_PDF.pdf](https://www.pnas.org/content/suppl/2008/09/11/0806162105.DCSupplemental/ST1_PDF.pdf): 
Homo sapiens (human), use 'hsa'; Mus musculus (mouse), use 'mmu', and Rattus norvegicus (rat), use 'rno'. The second,
'path' is also passed as a character string, and is the path of the directory in which the user would like to
write the data sets downloaded from the KEGG REST API. If the user does not provide a path, the data sets will 
be automatically written to the current working directory using the *here::here()* functionality. These data 
sets will be tab delimited files with a name describing the data, and for reproducibility, the date they were generated
and the version of KEGG when the API was accessed. In addition to these flat files, *get_kegg* will also
create a named list in R with the three relevant KEGG data sets. The names of this list will describe the data set.
For a detailed description of list elements use *?get_kegg*.

@@

In this step we will use the *pathEnrich* function to identify KEGG pathways that are enriched (i.e. over-represented) based on a gene list of interest. User gene lists must also be character vectors and be formatted as ENTREZ gene IDs. The *clusterProfiler* package offers a nice function (*bitr*) that maps gene symbols and Ensembl IDs to ENTREZ gene IDs, and an example can be seen in their [vignette](https://yulab-smu.github.io/clusterProfiler-book/chapter14.html#bitr).

@@

```{r pathEnrich_description, echo=FALSE}
# get column names
names <- colnames(list1_pe$enrich_table)
# Define column names
description <- c("KEGG Pathway Identifier",
                 "Description of KEGG Pathway (provided by KEGG)",
                 "Number of Genes in KEGG Pathway",
                 "Number of Genes from gene list in KEGG Pathway",
                 "Number of Genes in KEGG Database",
                 "Number of Genes from gene list in KEGG Database",
                 "Expected number of genes from list to be in KEGG pathway by chance",
                 "P-value for enrichment within the KEGG pathway for list genes",
                 "Multiple testing adjusted enrichment p-values (default = False Discovery Rate (Benjamini and Hochberg))",
                 "Ratio of number of genes observed from the gene list annotated to the KEGG pathway to the number of genes expected from the gene list to be annotated to the KEGG pathway if there was no enrichment (i.e. KEGG_PATHWAY_in_list/expected)")
# Make dataframe
df1 <- data.frame(names, description)
colnames(df1) <- c("Column Names", "Column Description") 
```

@@

```{r diffEnrich_descriptions, echo=FALSE}
## get vector of column names
de_col_names <- names(diff_enrich$de_table)
de_col_descr <- c("KEGG Pathway Identifier",
                  "Description of KEGG Pathway (provided by KEGG)",
                  "Number of Genes in KEGG Pathway",
                  "Number of Genes in KEGG Database",
                  "Number of Genes from gene list 1 in KEGG Pathway",
                  "Number of Genes from gene list 1 in KEGG Database",
                  "Expected number of genes from list 1 to be in KEGG pathway by chance",
                  "P-value for enrichment of list 1 genes related to KEGG pathway",
                  "Multiple testing adjusted enrichment p-values from gene list 1 (default = False Discovery Rate (Benjamini and Hochberg))",
                  "Ratio of number of genes observed from gene list 1 annotated to the KEGG pathway to the number of genes expected from gene list 1 annotated to the KEGG pathway if there was no enrichment (i.e. KEGG_PATHWAY_in_list1/expected_list1)",
                  "Number of Genes from gene list 2 in KEGG Pathway",
                  "Number of Genes from gene list 2 in KEGG Database",
                  "Expected number of genes from list 2 to be in KEGG pathway by chance",
                  "P-value for enrichment of list 2 genes related to KEGG pathway",
                  "Multiple testing adjusted enrichment p-values from gene list 2 (default = False Discovery Rate (Benjamini and Hochberg))",
                  "Ratio of number of genes observed from gene list 2 annotated to the KEGG pathway to the number of genes expected from gene list 2 annotated to the KEGG pathway if there was no enrichment (i.e. KEGG_PATHWAY_in_list2/expected_list2)",
                  "Odds of a gene from list 2 being from this KEGG pathway / Odds of a gene from list 1 being from this KEGG pathway",
                  "P-value for differential enrichment of this KEGG pathway between list 1 and list 2",
                  "Multiple testing adjusted differential enrichment p-values (default = False Discovery Rate (Benjamini and Hochberg))")
# Make dataframe
de_descr_df <- data.frame(de_col_names, de_col_descr)
colnames(de_descr_df) <- c("Column Names", "Column Description") 
```

@@

The result of the *diffEnrich* call is a list object that contains a data frame with the estimated odds ratio generated by the Fisher's Exact test 
and the associated p-value as well as other metadata from the function call. S3 generic functions for *print* and *summary* are
provided. The *print* function prints the results table as a *tibble*, and the
*summary* function returns the number of pathways that reached statistical significance
as well as their descriptions, the number of genes used from the KEGG database, the KEGG species,
the number of pathways that were shared (i.e. had at least 1 gene from each gene list present in the pathway) between the gene lists and the
method used for multiple testing correction.

@@

```{r echo=FALSE}
cap_plot <- c("Figure 2. Fold enrichment stratified by gene list. KEGG pathways are plotted on the y-axis and fold
enrichment is plotted on the x-axis. Each KEGG pathway has a bar depicting
its fold enrichment in list 1 (red) and its fold enrichment in list 2 (blue).
The transparency of the bars correspond to the unadjusted p-value for the
pathway's enrichment in the given list. The p-value presented as text to the
right of each pair of bars is the adjusted p-value (user defined: default is FDR) associated with the
differential enrichment of the pathway between the two lists, and the pathways
are ordered from top to bottom by this p-value (i.e. smallest p-value on top
of plot, and largest p-value on bottom of plot). The dotted line represents a fold enrichment of 1. Finally, the number of genes used
for analysis from each gene list (recall that this number may not be the same as the number of
genes in the user's original list) are reported below their respective p-values
in the legend.") 
```

@@

**Figure 2.** Fold enrichment stratified by gene list. KEGG pathways are plotted on the y-axis and fold
enrichment is plotted on the x-axis. Each KEGG pathway has a bar depicting
its fold enrichment in list 1 (red) and its fold enrichment in list 2 (blue).
The transparency of the bars correspond to the unadjusted p-value for the
pathway's enrichment in the given list. The p-value presented as text to the
right of each pair of bars is the adjusted p-value (user defined: default is FDR) associated with the
differential enrichment of the pathway between the two lists, and the pathways
are ordered from top to bottom by this p-value (i.e. smallest p-value on top
of plot, and largest p-value on bottom of plot). The dotted line represents a fold enrichment of 1. Finally, the number of genes used
for analysis from each gene list (recall that this number may not be the same as the number of
genes in the user's original list) are reported below their respective p-values
in the legend.

>>>>

In 2012 KEGG released its first application programming interface (API), and has been
adding features and functionality ever since. There are benefits to using an API.
First, API's, like KEGG's, allow users to perform customized analyses with the most
up-to-date versions of the data contained in the database. In addition, accessing the
KEGG API is very easy using statistical programming tools like R or Python and
integrating data retrieval into user's code makes the program reproducible. To 
further enforce reproducibilty *diffEnrich* adds a date and KEGG release tag to
all data files it generates from accessing the API. For update histories and release 
notes for the KEGG REST API please go [here](https://www.kegg.jp/kegg/rest/).

@@

**Figure 1. diffEnrich Analysis pipeline.** Functions within the diffEnrich package are represented by blue rectangles. The data that must be provided by the user is represented by the purple ovals. Data objects generated by a function in diffEnrich are represented by red ovals. The external call of the get_kegg function to the KEGG REST API is represented in yellow.

@@

Often high throughput omics studies include a functional enrichment analysis to glean biological insight from a list of candidate genes, proteins, metabolites, etc. Functional enrichment examines whether the number of genes in the list associated with a biological function or particular pathway is more than would be expected by chance.  As an example, enrichment of a particular pathway among a list of genes that are differentially expressed after an experimental manipulation may indicate that the pathway has been altered by that manipulation. This analysis is rather straight forward and many solutions have been offered (e.g., [Haung et al., 2009](https://www.ncbi.nlm.nih.gov/pubmed/19033363); [Kuleshov et al., 2016](https://www.ncbi.nlm.nih.gov/pubmed/27141961); [Liao et al., 2019](https://www.ncbi.nlm.nih.gov/pubmed/31114916); [Subramanian et al., 2005](https://www.pnas.org/content/102/43/15545)). A wide variety of databases have also been used to define these pathways (e.g., [Kanehisa and Goto, 2000](https://www.ncbi.nlm.nih.gov/pubmed/10592173)) and ontologies (e.g., [Ashburn et al., 2000](https://www.ncbi.nlm.nih.gov/pubmed/10802651)).

@@

### Step 1: Collect and store pathways from KEGG API
First we will use the *get_kegg* function to access the KEGG REST API and download the data sets required to perform 
our downstream analysis. This function takes two arguments. The first, 'species' is required. Currently, *diffEnrich* supports three species, and the argument is a character string using the KEGG
code [https://www.pnas.org/content/suppl/2008/09/11/0806162105.DCSupplemental/ST1_PDF.pdf](https://www.pnas.org/content/suppl/2008/09/11/0806162105.DCSupplemental/ST1_PDF.pdf): 
Homo sapiens (human), use 'hsa'; Mus musculus (mouse), use 'mmu'; and Rattus norvegicus (rat), use 'rno'. The second,
'path' is also passed as a character string, and is the path of the directory in which the user would like to
write the data sets downloaded from the KEGG REST API. If the user does not provide a path, the data sets will 
be automatically written to the current working directory using the *here::here()* functionality. These data 
sets will be tab delimited files with a name describing the data, and for reproducibility, the date they were generated
and the version of KEGG when the API was accessed. In addition to these flat files, *get_kegg* will also
create a named list in R with the three relevant KEGG data sets. The names of this list will describe the data set.
For a detailed description of list elements use *?get_kegg*.

@@

In this step we will use the *pathEnrich* function to identify KEGG pathways that are enriched (i.e. over-represented) based on a gene list of interest. User gene lists must be character vectors and be formatted as ENTREZ gene IDs. The *clusterProfiler* package offers a nice function (*bitr*) that maps gene symbols and Ensembl IDs to ENTREZ gene IDs, and an example can be seen in their [vignette](https://yulab-smu.github.io/clusterProfiler-book/chapter14.html#bitr).

@@

```{r pathEnrich_description, echo=FALSE}
# get column names
names <- colnames(list1_pe$enrich_table)
# Define column names
description <- c("KEGG Pathway Identifier",
                 "Description of KEGG Pathway (provided by KEGG)",
                 "Number of Genes in KEGG Pathway",
                 "Number of Genes from gene list in KEGG Pathway",
                 "Number of Genes in KEGG Database",
                 "Number of Genes from gene list in KEGG Database",
                 "Expected number of genes from list to be in KEGG pathway by chance",
                 "P-value for enrichment within the KEGG pathway for list genes",
                 "Multiple testing adjusted enrichment p-values (default = False Discovery Rate (Benjamini and Hochberg, 1995))",
                 "Ratio of number of genes observed from the gene list annotated to the KEGG pathway to the number of genes expected from the gene list to be annotated to the KEGG pathway if there was no enrichment (i.e. KEGG_PATHWAY_in_list/expected)")
# Make dataframe
df1 <- data.frame(names, description)
colnames(df1) <- c("Column Names", "Column Description") 
```

@@

```{r diffEnrich_descriptions, echo=FALSE}
## get vector of column names
de_col_names <- names(diff_enrich$de_table)
de_col_descr <- c("KEGG Pathway Identifier",
                  "Description of KEGG Pathway (provided by KEGG)",
                  "Number of Genes in KEGG Pathway",
                  "Number of Genes in KEGG Database",
                  "Number of Genes from gene list 1 in KEGG Pathway",
                  "Number of Genes from gene list 1 in KEGG Database",
                  "Expected number of genes from list 1 to be in KEGG pathway by chance",
                  "P-value for enrichment of list 1 genes related to KEGG pathway",
                  "Multiple testing adjusted enrichment p-values from gene list 1 (default = False Discovery Rate (Benjamini and Hochberg, 1995))",
                  "Ratio of number of genes observed from gene list 1 annotated to the KEGG pathway to the number of genes expected from gene list 1 annotated to the KEGG pathway if there was no enrichment (i.e. KEGG_PATHWAY_in_list1/expected_list1)",
                  "Number of Genes from gene list 2 in KEGG Pathway",
                  "Number of Genes from gene list 2 in KEGG Database",
                  "Expected number of genes from list 2 to be in KEGG pathway by chance",
                  "P-value for enrichment of list 2 genes related to KEGG pathway",
                  "Multiple testing adjusted enrichment p-values from gene list 2 (default = False Discovery Rate (Benjamini and Hochberg, 1995))",
                  "Ratio of number of genes observed from gene list 2 annotated to the KEGG pathway to the number of genes expected from gene list 2 annotated to the KEGG pathway if there was no enrichment (i.e. KEGG_PATHWAY_in_list2/expected_list2)",
                  "Odds of a gene from list 2 being from this KEGG pathway / Odds of a gene from list 1 being from this KEGG pathway",
                  "P-value for differential enrichment of this KEGG pathway between list 1 and list 2",
                  "Multiple testing adjusted differential enrichment p-values (default = False Discovery Rate (Benjamini and Hochberg, 1995))")
# Make dataframe
de_descr_df <- data.frame(de_col_names, de_col_descr)
colnames(de_descr_df) <- c("Column Names", "Column Description") 
```

@@

The result of the *diffEnrich* call is a list object that contains a data frame with the estimated odds ratio generated by the Fisher's Exact test 
and the associated p-value as well as other metadata from the function call. S3 generic functions for *print* and *summary* are
provided. The *print* function prints the results table as a *tibble*, and the
*summary* function returns the number of pathways that reached statistical significance
as well as their descriptions, the number of genes used from the KEGG database, the KEGG species,
the number of pathways that were shared (i.e. had at least N gene from each gene list present in the pathway based on what the user chose for N in *pathEnrich*) between the gene lists and the
method used for multiple testing correction.

@@

```{r echo=FALSE}
cap_plot <- c("Figure 2. Example of a differential enrichment graphic. KEGG pathways are plotted on the y-axis and fold
enrichment is plotted on the x-axis. Each KEGG pathway has a bar depicting
its fold enrichment in list 1 (red) and its fold enrichment in list 2 (blue).
The transparency of the bars correspond to the unadjusted p-value for the
pathway's enrichment in the given list. The p-value presented as text to the
right of each pair of bars is the adjusted p-value (user defined: default is FDR) associated with the
differential enrichment of the pathway between the two lists, and the pathways
are ordered from top to bottom by this p-value (i.e. smallest p-value on top
of plot, and largest p-value on bottom of plot). The dotted line represents a fold enrichment of 1. Finally, the number of genes used
for analysis from each gene list (recall that this number may not be the same as the number of
genes in the user's original list) are reported below their respective p-values
in the legend.") 
```

@@

**Figure 2.  Example of a differential enrichment graphic.** KEGG pathways are plotted on the y-axis and fold
enrichment is plotted on the x-axis. Each KEGG pathway has a bar depicting
its fold enrichment in list 1 (red) and its fold enrichment in list 2 (blue).
The transparency of the bars correspond to the unadjusted p-value for the
pathway's enrichment in the given list. The p-value presented as text to the
right of each pair of bars is the adjusted p-value (user defined: default is FDR) associated with the
differential enrichment of the pathway between the two lists, and the pathways
are ordered from top to bottom by this p-value (i.e. smallest p-value on top
of plot, and largest p-value on bottom of plot). The dotted line represents a fold enrichment of 1. Finally, the number of genes used
for analysis from each gene list (recall that this number may not be the same as the number of
genes in the user's original list) are reported below their respective p-values
in the legend.
